A hazards approach to the biometric analysis of infant mortality

### 2023-03-26 paper2.Rmd
add maps and time periods in figure text

Bourgeois-Pichat also found deviations from the linear death pattern after the first month, the curve for Sardinia in 1948 bend strongly upward after four months [@bp52, Figure 12]. @knoki77 give ample examples of upward bending curves and argue that they are due to increasing mortality after stopping breast feeding. In Quebec 1944--1947, the curve bends down quite strongly after the sixth month [@bp52, Figure 8]. A downward bend is also what has been found using historical Swedish parish data [@tb99, Figure 6b; @tbcl99, Figures 6 and 7; @lyn98, Figures 1 and 2; @jslgt81, Figure 6d], though less pronounced than in Quebec. The curve for Sweden 1910--1946 show, however, no such downward bend [@bp51b, Figure 9; @bp52].

@@

```{r logcube, fig.cap = "The log-cube transform of time in days versus the identity transform (dashed). Note that g(365) = 365.", fig.height = 7}
x <- seq(0, 365, length = 366)
plot(x[1:4], g(x[1:4]), type = "l", xlab = "Days", ylab = "g(Days)", 
     col = "blue", axes = FALSE)
att <- c(0, 28, 183, 365)
axis(1, at = att, labels = att)
axis(2, las = 1, at = g(att), labels = c("0", "g(28)", "g(183)", "365"))
box()
lines(c(0, 365), c(0, 365), lty = 2)
abline(v = c(0, 365), h = c(0, 365))
abline(v = 28, h = g(28), col = "darkgreen", lty = 3)
```

@@

Assume that a cohort of infants is followed over time from birth to age one. There are no drop-outs 
(no right censoring except at age 365 days). The exact age in days at each observed death is noted and transformed by $g$, 
and due to imperfect time measuring there are often tied death ages. As an illustrative example, 
we use a data set from northern Sweden, covering the years 1861--1950.
```{r readdata}
infant <- readRDS("Rdata/infant.rds")
infant$month <- months(infant$birthdate)
infant$period <- cut(infant$birthdate, 
                   breaks = as.Date(c("1860-12-31", "1890-12-31", 
                                      "1920-12-31", "1951-01-01")),
                   labels = c("1861-1890", "1891-1920", "1921-1950"))
infant <- infant[, c("birthdate", "sex", "enter", "exit", "event", "period", "month")]
ginfant <- infant
ginfant$enter <- g(infant$enter)
ginfant$exit <- g(infant$exit)
```
For the first and last 30 years in this data set, we have the results shown in Figure \@ref(fig:first).

@@

```{r third, fig.cap = "The hazards method plotted in the Bourgeous-Pichat way."}
oldpar <- par(mfrow = c(1, 2))
drawBB2(g1, main = "Type hazards plot", ylim = c(0, 0.12))
source("R/drawBB.R")
drawBB(r1, main = "Type Bourgeois-Pichat plot", ylim = c(0, 0.12))
par(oldpar)
```

The left hand panel is clearly more informative, we see both the exogenous and the endogenous
cumulative plots. The dashed blue line in the left hand panel is the exogenous cumulative hazard function,
while it in the right hand panel is the same function, but shifted upwards 0.027 units.
It is a consequence of the "paper, pencil and ruler" method used by @bp52a, and it can be
argued that it gives a visual perception of goodness-of-fit lacking in the left hand panel. 
But that is not relevant, because the same perception of goodness-of fit is achieved by
looking at the estimated endogenous cumulative hazard function: It should be a horizontal straight 
line in the post-neonatal age period.

@@

```{r enof, fig.cap = "Cumulative hazard functions for neonatal mortality, Västerbotten.", fig.width = 7}
yen <- y - 1000 * rate * x ##"Endo"
par(las = 1)
plot(x, yen, type = "s", col = "red", xlab = "Day", ylab = "by 1000", ylim = c(0, 36), axes = FALSE)
axis(2, las = 1)
axis(1, at = c(g(0), g(7), g(14), g(21), g(28)), labels = c(0, 7, 14, 21, 28))
box()
abline(h = 0, v = 0)
lines(x, 1000 * rate * x, col = "blue", lty = 3)
lines(x, y, lty = 1, type = "s")
text(8, 20, "Total", cex = 0.8)
text(15, 14, "Endogenous", col = "red")
text(8, 5, "Exogenous", col = "blue", cex = 0.8)
n <- length(yen)
axis(4, at = c(0, 1000 * rate * x[n], yen[n], y[n]), las = 1, cex.axis = 0.7)
```

@@

```{r plotcohort, fig.cap = "By birth month, cohorts, Västerbotten 1861-1950.", fig.height = 8, eval = TRUE}
oldpar <- par(mfrow = c(1, 1))
dato <- dat[dat$birthdate < as.Date("1950-12-31"), ]
fit <- coxreg(Surv(g(enter), g(exit), event) ~ strata(bmonth), data = dato)
cols <- c("blue", "magenta", rep("red", 4), rep("green", 3), "magenta", "blue", "blue")
haz <- hazards(fit)
pd <- numeric(12)
for(i in 1:12){
    haz[[i]] <- rbind(c(0, 0), haz[[i]])
    haz[[i]][, 2] <- cumsum(haz[[i]][, 2])
    pd[i] <- round((1 - exp(-max(haz[[i]][, 2]))) * 100, 1)
}
plot(haz[[1]][, 1], haz[[1]][, 2], col = cols[1], axes = FALSE, 
     type = "l", xlab = "Days", ylab = "Cumulative hazards")
for (i in 2:12){
    lines(haz[[i]][, 1], haz[[i]][, 2], col = cols[i], type = "l", lty = i)
}
legend("topleft", legend = paste(month.abb, pd, sep = ":  "), lty = 1:12, col = cols)
att <- c(0, 28, 183, 365)
axis(1, at = g(att), labels = att)
axis(2, las = 1)
box()
abline(v = g(28), lty = 3)
abline(h = 0)
fita <- coxreg(Surv(g(enter), g(exit), event) ~ 1, data = dato)
haza <- hazards(fita)
lines(c(0, haza[[1]][, 1]), cumsum(c(0, haza[[1]][, 2])), type = "l", lwd = 2, col = "black")
legend(100, 0.08, legend = "All", lty = 1, col = "black", lwd = 2)
par(oldpar)
```

@@

```{r plotcohort3, fig.cap = "By month, period data, Västerbotten 1861-1950.", fig.height = 8, eval = TRUE}
oldpar <- par(mfrow = c(1, 1))
dato <- dat[dat$birthdate < as.Date("1950-12-31"), ]
fit <- coxreg(Surv(g(enter), g(exit), event) ~ strata(month), data = dato)
cols <- c("blue", "magenta", rep("red", 4), rep("green", 3), "magenta", "blue", "blue")
haz <- hazards(fit)
pd <- numeric(12)
for(i in 1:12){
    haz[[i]] <- rbind(c(0, 0), haz[[i]])
    haz[[i]][, 2] <- cumsum(haz[[i]][, 2])
    pd[i] <- round((1 - exp(-max(haz[[i]][, 2]))) * 100, 1)
}
plot(haz[[1]][, 1], haz[[1]][, 2], col = cols[1], axes = FALSE, 
     type = "l", xlab = "Days", ylab = "Cumulative hazards")
for (i in 2:12){
    lines(haz[[i]][, 1], haz[[i]][, 2], col = cols[i], type = "l", lty = i)
}
legend("topleft", legend = paste(month.abb, pd, sep = ":  "), lty = 1:12, col = cols)
att <- c(0, 28, 183, 365)
axis(1, at = g(att), labels = att)
axis(2, las = 1)
box()
abline(v = g(28), lty = 3)
abline(h = 0)
fita <- coxreg(Surv(g(enter), g(exit), event) ~ 1, data = dato)
haza <- hazards(fita)
lines(c(0, haza[[1]][, 1]), cumsum(c(0, haza[[1]][, 2])), type = "l", lwd = 2, col = "black")
legend(100, 0.08, legend = "All", lty = 1, col = "black", lwd = 2)
par(oldpar)
```

```{r plotperiod, fig.cap = "By calendar month, period data, Västerbotten 1861-1950.", fig.height = 9, eval = FALSE}
oldpar <- par(mfrow = c(4, 3))
for (mon in levels(dat$bmonth)){
    drawBB2(dat[dat$month == mon, ], main = mon, printLegend = FALSE)
}
par(oldpar)
```

@@

## Scania 1710--1800.

The Scanian dataset covers five parishes in Scania 1710--1800 [@sedd].
It is an interesting case because it appears as if the data quality is low since 
the linearity during the post-neonatal period is gone. Rather, there is a "hockey-stick" pattern, 
also seen in other old data sets. See Figure \@ref(fig:scahigh).

@@

```{r scahigh, fig.cap = "Scania 1710--1800."}
sca1 <- readRDS("Rdata/sca1.rds")
sca1 <- sca1[sca1$BirthDate < "1801-01-01", ]
sca11 <- sca1
sca1$exit <- g(sca1$exit)
sca1$enter <- g(sca1$enter)
rs <- with(sca1, risksets(Surv(enter, exit, event), members = FALSE))
oldpar <- par(mfrow = c(1, 2))
drawBB2(sca11, main = "Hazards method")
drawBP(rs, main = "B-P method")
par(oldpar)
```

>>>>

Bourgeois-Pichat also found deviations from the linear death pattern after the first month, the curve for Sardinia in 1948 bend strongly upward after four months [@bp52, Figure 12]. @knoki77 give ample examples of upward bending curves and argue that they are due to increasing mortality after stopping breast feeding. In Quebec 1944--1947, the curve bends down quite strongly after the sixth month [@bp52, Figure 8]. A downward bend is also what has been found using historical Swedish parish data [@tb99, Figure 6b; @tbcl99, Figures 6 and 7; @lyn98, Figures 1 and 2; @jslgt81, Figure 6d], though less pronounced than in Quebec. The curve for Sweden 1910--1946 show, however, no such downward bend but a straight line [@bp51b, Figure 9; @bp52].

@@

```{r logcube, fig.cap = "The log-cube transform of time in days versus the identity transform (dashed). Note that g(365) = 365.", fig.scap = "The log-cube transform", fig.height = 7}
x <- seq(0, 365, length = 366)
plot(x, g(x), type = "l", xlab = "Days", ylab = "g(Days)", 
     col = "blue", axes = FALSE)
att <- c(0, 28, 183, 365)
axis(1, at = att, labels = att)
axis(2, las = 1, at = g(att), labels = c("0", "g(28)", "g(183)", "365"))
box()
lines(c(0, 365), c(0, 365), lty = 2)
abline(v = c(0, 365), h = c(0, 365))
abline(v = 28, h = g(28), col = "darkgreen", lty = 3)
```

@@

Assume that a cohort of infants is followed over time from birth to age one. There are no drop-outs 
(no right censoring except at age 365 days). The exact age in days at each observed death is noted and transformed by $g$, 
and due to imperfect time measuring there are often tied death ages. As an illustrative example, 
we use a data set from Västerbotten in northern Sweden, covering the years 1861--1950, see 
Figure \@ref(fig:vbmap).

```{r vbmap, fig.cap="Västerbotten."}
knitr::include_graphics("Slides/myfigs/vb.png")
```

```{r readdata}
infant <- readRDS("Rdata/infant.rds")
infant$month <- months(infant$birthdate)
infant$period <- cut(infant$birthdate, 
                   breaks = as.Date(c("1860-12-31", "1890-12-31", 
                                      "1920-12-31", "1951-01-01")),
                   labels = c("1861-1890", "1891-1920", "1921-1950"))
infant <- infant[, c("birthdate", "sex", "enter", "exit", "event", "period", "month")]
ginfant <- infant
ginfant$enter <- g(infant$enter)
ginfant$exit <- g(infant$exit)
```
For the first and last 30 years in this data set, we have the results shown in Figure \@ref(fig:first).

@@

```{r third, fig.cap = "The hazards method plotted in the Bourgeous-Pichat way, Västerbotten 1861--1890."}
oldpar <- par(mfrow = c(1, 2))
drawBB2(g1, main = "Type hazards plot", ylim = c(0, 0.12))
source("R/drawBB.R")
drawBB(r1, main = "Type Bourgeois-Pichat plot", ylim = c(0, 0.12))
par(oldpar)
```

The left hand panel is clearly more informative, we see both the exogenous and the endogenous
cumulative plots. The dashed blue line in the left hand panel is the exogenous cumulative hazard function,
while it in the right hand panel is the same function, but shifted upwards 0.027 units.
It is a consequence of the "paper, pencil and ruler" method used by @bp51a, and it can be
argued that it gives a visual perception of goodness-of-fit lacking in the left hand panel. 
But that is not relevant, because the same perception of goodness-of fit is achieved by
looking at the estimated endogenous cumulative hazard function: It should be a horizontal straight 
line in the post-neonatal age period.

@@

```{r enof, fig.cap = "Cumulative hazard functions for neonatal mortality, Västerbotten 1861--1950.", fig.width = 7}
yen <- y - 1000 * rate * x ##"Endo"
par(las = 1)
plot(x, yen, type = "s", col = "red", xlab = "Day", ylab = "by 1000", ylim = c(0, 36), axes = FALSE)
axis(2, las = 1)
axis(1, at = c(g(0), g(7), g(14), g(21), g(28)), labels = c(0, 7, 14, 21, 28))
box()
abline(h = 0, v = 0)
lines(x, 1000 * rate * x, col = "blue", lty = 3)
lines(x, y, lty = 1, type = "s")
text(8, 20, "Total", cex = 0.8)
text(15, 14, "Endogenous", col = "red")
text(8, 5, "Exogenous", col = "blue", cex = 0.8)
n <- length(yen)
axis(4, at = c(0, 1000 * rate * x[n], yen[n], y[n]), las = 1, cex.axis = 0.7)
```

@@

```{r plotcohort, fig.cap = "By birth month, cohorts, Västerbotten 1861--1950.", fig.height = 8, eval = TRUE}
oldpar <- par(mfrow = c(1, 1))
dato <- dat[dat$birthdate < as.Date("1950-12-31"), ]
fit <- coxreg(Surv(g(enter), g(exit), event) ~ strata(bmonth), data = dato)
cols <- c("blue", "magenta", rep("red", 4), rep("green", 3), "magenta", "blue", "blue")
haz <- hazards(fit)
pd <- numeric(12)
for(i in 1:12){
    haz[[i]] <- rbind(c(0, 0), haz[[i]])
    haz[[i]][, 2] <- cumsum(haz[[i]][, 2])
    pd[i] <- round((1 - exp(-max(haz[[i]][, 2]))) * 100, 1)
}
plot(haz[[1]][, 1], haz[[1]][, 2], col = cols[1], axes = FALSE, 
     type = "l", xlab = "Days", ylab = "Cumulative hazards")
for (i in 2:12){
    lines(haz[[i]][, 1], haz[[i]][, 2], col = cols[i], type = "l", lty = i)
}
legend("topleft", legend = paste(month.abb, pd, sep = ":  "), lty = 1:12, col = cols)
att <- c(0, 28, 183, 365)
axis(1, at = g(att), labels = att)
axis(2, las = 1)
box()
abline(v = g(28), lty = 3)
abline(h = 0)
fita <- coxreg(Surv(g(enter), g(exit), event) ~ 1, data = dato)
haza <- hazards(fita)
lines(c(0, haza[[1]][, 1]), cumsum(c(0, haza[[1]][, 2])), type = "l", lwd = 2, col = "black")
legend(100, 0.08, legend = "All", lty = 1, col = "black", lwd = 2)
par(oldpar)
```

@@

```{r plotcohort3, fig.cap = "By month, period data, Västerbotten 1861--1950.", fig.height = 8, eval = TRUE}
oldpar <- par(mfrow = c(1, 1))
dato <- dat[dat$birthdate < as.Date("1950-12-31"), ]
fit <- coxreg(Surv(g(enter), g(exit), event) ~ strata(month), data = dato)
cols <- c("blue", "magenta", rep("red", 4), rep("green", 3), "magenta", "blue", "blue")
haz <- hazards(fit)
pd <- numeric(12)
for(i in 1:12){
    haz[[i]] <- rbind(c(0, 0), haz[[i]])
    haz[[i]][, 2] <- cumsum(haz[[i]][, 2])
    pd[i] <- round((1 - exp(-max(haz[[i]][, 2]))) * 100, 1)
}
plot(haz[[1]][, 1], haz[[1]][, 2], col = cols[1], axes = FALSE, 
     type = "l", xlab = "Days", ylab = "Cumulative hazards")
for (i in 2:12){
    lines(haz[[i]][, 1], haz[[i]][, 2], col = cols[i], type = "l", lty = i)
}
legend("topleft", legend = paste(month.abb, pd, sep = ":  "), lty = 1:12, col = cols)
att <- c(0, 28, 183, 365)
axis(1, at = g(att), labels = att)
axis(2, las = 1)
box()
abline(v = g(28), lty = 3)
abline(h = 0)
fita <- coxreg(Surv(g(enter), g(exit), event) ~ 1, data = dato)
haza <- hazards(fita)
lines(c(0, haza[[1]][, 1]), cumsum(c(0, haza[[1]][, 2])), type = "l", lwd = 2, col = "black")
legend(100, 0.08, legend = "All", lty = 1, col = "black", lwd = 2)
par(oldpar)
```

```{r plotperiod, fig.cap = "By calendar month, period data, Västerbotten 1861--1950.", fig.height = 9, eval = FALSE}
oldpar <- par(mfrow = c(4, 3))
for (mon in levels(dat$bmonth)){
    drawBB2(dat[dat$month == mon, ], main = mon, printLegend = FALSE)
}
par(oldpar)
```

@@

## Skåne 1710--1800.

Skåne is the southernmost province of Sweden, see the map in Figure \@ref(fig:scaniamap).

```{r scaniamap, fig.cap = "Skåne."}
knitr::include_graphics("Slides/myfigs/scania.png")
```

This dataset covers five parishes i n Skåne 1710--1800 [@sedd].
It is an interesting case because it appears as if the data quality is low since 
the linearity during the post-neonatal period is gone. Rather, there is a "hockey-stick" pattern, 
also seen in other old data sets. See Figure \@ref(fig:scahigh).

@@

```{r scahigh, fig.cap = "Skåne 1710--1800."}
sca1 <- readRDS("Rdata/sca1.rds")
sca1 <- sca1[sca1$BirthDate < "1801-01-01", ]
sca11 <- sca1
sca1$exit <- g(sca1$exit)
sca1$enter <- g(sca1$enter)
rs <- with(sca1, risksets(Surv(enter, exit, event), members = FALSE))
oldpar <- par(mfrow = c(1, 2))
drawBB2(sca11, main = "Hazards method")
drawBP(rs, main = "B-P method")
par(oldpar)
```
